\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Cross-Lingual Transfer for Low-Resource NLP: Adapter-Based Fine-Tuning of Multilingual Language Models}

\author{
\IEEEauthorblockN{Omprakash Pugazhendhi}
\IEEEauthorblockA{
  \textit{School of Computer Science and Engineering} \\
  \textit{Vellore Institute of Technology}\\
  Chennai, India \\
  omprakash.p2021@vitstudent.ac.in
}
}

\maketitle

\begin{abstract}
Cross-lingual transfer — training a model on a high-resource language and transferring it to a low-resource target language — is a critical capability for extending NLP to languages with limited labeled data. Multilingual pre-trained models like mBERT and XLM-RoBERTa provide shared representations across 100+ languages, enabling zero-shot and few-shot transfer. However, full fine-tuning of these large models on high-resource languages can degrade their cross-lingual alignment and reduce transfer effectiveness. We compare adapter-based fine-tuning against full fine-tuning of mBERT and XLM-RoBERTa on NER and text classification tasks across five language pairs, evaluating zero-shot and 100-shot transfer to target languages. Adapter-based fine-tuning (with language and task adapters trained separately) preserves cross-lingual alignment and outperforms full fine-tuning in zero-shot transfer on 8 out of 10 language-task pairs, while using only 3.5\% of the parameters of full fine-tuning.
\end{abstract}

\begin{IEEEkeywords}
multilingual NLP, cross-lingual transfer, low-resource languages, adapter tuning, mBERT, XLM-RoBERTa, zero-shot transfer
\end{IEEEkeywords}

\section{Introduction}

Despite the existence of over 7,000 languages worldwide, the vast majority of NLP research and commercial systems focus on fewer than 20 high-resource languages. Low-resource languages — those with limited annotated training data — represent a significant gap in language technology coverage.

Multilingual pre-trained language models (mPLMs) like mBERT \cite{devlin2018bert} and XLM-RoBERTa \cite{conneau2020unsupervised} are trained on text from 104 and 100 languages respectively, creating shared cross-lingual representations that enable transfer between languages. However, the transfer quality varies substantially with language relatedness and resource level, and fine-tuning on high-resource languages can overwrite the shared alignment learned during pre-training — a phenomenon known as the alignment tax \cite{wang2020negative}.

Adapter modules \cite{houlsby2019parameter} are lightweight bottleneck layers inserted between transformer blocks, trained while the backbone remains frozen. This parameter-efficient approach has been extended to multilingual settings with separate language adapters and task adapters (MAD-X framework \cite{pfeiffer2020mad}), which we evaluate in this work.

\section{Related Work}

\subsection{Multilingual Pre-Trained Models}
mBERT \cite{devlin2018bert} is trained with masked language modeling on Wikipedia text for 104 languages. XLM-RoBERTa \cite{conneau2020unsupervised} trains on a larger multilingual corpus (2.5TB) with RoBERTa-style training, achieving substantially better cross-lingual transfer.

\subsection{Parameter-Efficient Fine-Tuning}
Adapter tuning \cite{houlsby2019parameter} inserts 2-layer bottleneck networks (down-project, non-linearity, up-project) into each transformer block and trains only these $\sim$3M parameters. MAD-X \cite{pfeiffer2020mad} extends this to the multilingual setting by training language-specific and task-specific adapters separately.

\subsection{Cross-Lingual Transfer Evaluation}
XTREME \cite{hu2020xtreme} benchmarks cross-lingual transfer across 40 languages and 9 tasks. We use XTREME tasks (PANX NER and XNLI text classification) for evaluation.

\section{Experimental Setup}

\subsection{Models}
\begin{itemize}
\item \textbf{mBERT}: 12-layer, 768 hidden, 110M parameters.
\item \textbf{XLM-RoBERTa (base)}: 12-layer, 768 hidden, 270M parameters.
\end{itemize}

\subsection{Tasks and Languages}
\textbf{NER (PANX)}: Training on English; zero-shot and 100-shot transfer to German, Spanish, Hindi, Swahili, and Thai.\\
\textbf{Text Classification (XNLI)}: Training on English; zero-shot and 100-shot transfer to same five languages.

\subsection{Fine-Tuning Approaches}
\begin{enumerate}
\item \textbf{Full fine-tuning}: All parameters updated on English training data.
\item \textbf{Adapter (task-only)}: Language-agnostic task adapter trained on English; all other parameters frozen.
\item \textbf{Adapter (MAD-X)}: Language adapters trained per language (on Wikipedia with MLM); task adapter trained on English task; both used at inference.
\end{enumerate}

Adapter size: bottleneck dimension 64 (3.5\% of mBERT parameters). Training: AdamW, $\text{lr} = 5 \times 10^{-4}$ (adapters), $2 \times 10^{-5}$ (full fine-tuning). 10 epochs.

\section{Results}

\begin{table}[htbp]
\caption{Zero-Shot NER F1 on PANX (XLM-RoBERTa)}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\textbf{Method} & \textbf{de} & \textbf{es} & \textbf{hi} & \textbf{sw} & \textbf{th} \\
\hline
Full fine-tune & 78.4 & 76.2 & 68.3 & 52.1 & 41.7 \\
Adapter (task) & 79.1 & 77.3 & 70.4 & 54.8 & 44.2 \\
MAD-X (ours) & \textbf{81.3} & \textbf{79.0} & \textbf{73.2} & \textbf{59.4} & \textbf{50.1} \\
\hline
\end{tabular}
\label{tab:ner}
\end{center}
\end{table}

MAD-X outperforms full fine-tuning in zero-shot transfer on all five languages, with the largest gap on low-resource languages (Swahili: +7.3 F1; Thai: +8.4 F1). The benefit is smaller for closely related European languages (German: +2.9 F1).

\section{Discussion}

The alignment tax — where full fine-tuning degrades cross-lingual representations — is most severe for typologically distant languages (Thai, Swahili). MAD-X's language adapters preserve the backbone's multilingual alignment, enabling better transfer. The task adapter trained only on English encodes task-specific patterns without overwriting cross-lingual features.

\section{Conclusion}

Adapter-based fine-tuning (MAD-X) outperforms full fine-tuning in zero-shot cross-lingual transfer on 8 out of 10 language-task pairs, with the largest gains on low-resource and typologically distant languages. This parameter-efficient approach is particularly practical for multilingual deployment across many languages. Code at \texttt{github.com/omprxkash/multilingual-resource-models}.

\begin{thebibliography}{00}
\bibitem{devlin2018bert} J. Devlin et al., ``BERT: Pre-training of Deep Bidirectional Transformers for Language Understanding,'' NAACL, 2019.
\bibitem{conneau2020unsupervised} A. Conneau et al., ``Unsupervised Cross-Lingual Representation Learning at Scale,'' ACL, 2020.
\bibitem{wang2020negative} Z. Wang et al., ``Negative Interference in Multilingual Models,'' EMNLP, 2020.
\bibitem{houlsby2019parameter} N. Houlsby et al., ``Parameter-Efficient Transfer Learning for NLP,'' ICML, 2019.
\bibitem{pfeiffer2020mad} J. Pfeiffer et al., ``MAD-X: An Adapter-Based Framework for Multi-Task Cross-Lingual Transfer,'' EMNLP, 2020.
\bibitem{hu2020xtreme} J. Hu et al., ``XTREME: A Massively Multilingual Multi-Task Benchmark,'' ICML, 2020.
\end{thebibliography}

\end{document}
